---
title: "Preliminary Analysis of World Happiness"
author: "Claire Gloss (cg4ey)"
date: "4/8/2022"
output: word_document
---
Data set: World Happiness Report
\
Question in mind: Best and Worst Countries/Regions to Live
```{r echo=F, message=F}
##load the data frame and the packages required for the code
df<-read.csv("data_all.csv", header=T)
#names(df) #check names of columns
library(tidyverse)
library(ggplot2)
```
Many variables in this data set are a proportion of contribution to the happiness ranking. So, the only variable that will be indicative of the best or worst places to live will be the happiness score. 
\
\
Let's look at the happiness scores for each year individually and rank the top five and bottom five countries.
\
```{r echo=F}
##separate the data frame by year
df2015<-df%>%
  filter(Year==2015)%>%
  arrange(desc(Happiness.Score))
df2016<-df%>%
  filter(Year==2016)%>%
  arrange(desc(Happiness.Score))
df2017<-df%>%
  filter(Year==2017)%>%
  arrange(desc(Happiness.Score))
df2018<-df%>%
  filter(Year==2018)%>%
  arrange(desc(Happiness.Score))
df2019<-df%>%
  filter(Year==2019)%>%
  arrange(desc(Happiness.Score))
df2020<-df%>%
  filter(Year==2020)%>%
  arrange(desc(Happiness.Score))
df2021<-df%>%
  filter(Year==2021)%>%
  arrange(desc(Happiness.Score))
df2022<-df%>%
  filter(Year==2022)%>%
  arrange(desc(Happiness.Score))
```
```{r echo=F}
##Top and Bottom 5 countries
BW2015<-rbind(head(df2015['Country'],5),tail(df2015['Country'],5))
BW2016<-rbind(head(df2016['Country'],5),tail(df2016['Country'],5))
BW2017<-rbind(head(df2017['Country'],5),tail(df2017['Country'],5))
BW2018<-rbind(head(df2018['Country'],5),tail(df2018['Country'],5))
BW2019<-rbind(head(df2019['Country'],5),tail(df2019['Country'],5))
BW2020<-rbind(head(df2020['Country'],5),tail(df2020['Country'],5))
BW2021<-rbind(head(df2021['Country'],5),tail(df2021['Country'],5))
BW2022<-rbind(head(df2022['Country'],5),tail(df2022['Country'],5))

##rename so we can combine into one data frame
names(BW2015)[1]<-'2015'
names(BW2016)[1]<-'2016'
names(BW2017)[1]<-'2017'
names(BW2018)[1]<-'2018'
names(BW2019)[1]<-'2019'
names(BW2020)[1]<-'2020'
names(BW2021)[1]<-'2021'
names(BW2022)[1]<-'2022'

##re-index so we can combine into one data frame and so it is logical
new_index<-c(1,2,3,4,5,-5,-4,-3,-2,-1)
rownames(BW2015)<-new_index
rownames(BW2016)<-new_index
rownames(BW2017)<-new_index
rownames(BW2018)<-new_index
rownames(BW2019)<-new_index
rownames(BW2020)<-new_index
rownames(BW2021)<-new_index
rownames(BW2022)<-new_index

##combine data frames to show top and worst countries across the years
BW<-data.frame(BW2015[1],BW2016[1],BW2017[1],BW2018[1],BW2019[1],BW2020[1],BW2021[1],BW2022[1])
names(BW)<- c('2015','2016','2017','2018','2019','2020','2021','2022')
print(BW)

```
\
\
The above charts shows that there are a lot of repeat countries across the years, so now we want to look now at the average scores across the years and see which countries have the happiest and least happy residents overall. 
\
```{r echo=F}
##create list of the countries in the data frame and find the average happiness scores for each
countries<-unique(df$Country)
scores<-c()
for (x in countries) {
  new_df<-df%>%
    filter(Country==x)
  mean<-mean(new_df$Happiness.Score,na.rm=T)
  scores<-c(scores,mean)
}

```
```{r echo=F}
##put into new data frame and rank the countries by the happiness score
df_avg<-data.frame(countries, scores)
df_avg<-df_avg%>%
  arrange(desc(scores))

##show top and bottom 5 countries and re-index for logical sense
BW_avg<-rbind(head(df_avg,5),tail(df_avg,5))
rownames(BW_avg)<-new_index #same index as before
print(BW_avg[1])
```
\
I am curious to see just how much more happy each country is than one another, so let's look at a plot to see the deltas.
\
```{r echo=F, warning=F}
par(mrow=c(1,2))
ggplot(data=head(BW_avg,5),aes(x=countries, y=scores))+
  geom_point()+
  labs(x='Country', y='Happiness Score', title='Happiest Countries')
ggplot(data=tail(BW_avg,5),aes(x=countries, y=scores))+
  geom_point()+
  labs(x='Country', y='Happiness Score', title='Least Happy Countries')
par(mrow=c(1,1))
```
\
These plots indicate that Finnish and Danish people are significantly happier than the other "happiest" people, and Afghans and Central Africans are significantly less happy than the other "least happy" people. For a better understanding of these differences, let's look at the percent differences in the happiness scores across these countries. 
\
```{r echo=F}
##Figure the percent difference between the happiest/least happy countries and the average happiness of all countries
mean_score<-mean(df_avg$scores)
BW_avg<-BW_avg%>%
  mutate(delta_to_mean=round(((scores/mean_score)-1)*100,2))
print(BW_avg[c(1,3)])

```
\
The above chart shows that Finnish people are 42.46% happier than other people and Afghans are 41.63% less happy than other people. 
\
\
Now, let's compare each country to the average score for the five "happiest" countries and average score for the five "least happy" countries.
\
```{r echo=F}
#Identify mean scores for the happiest nations and mean scores for the least happy nations
mean_happy<-mean(head(BW_avg,5)$scores)
mean_sad<-mean(tail(BW_avg,5)$scores)

#find the percent differences 
BW_avg<-BW_avg%>%
  mutate(delta_to_happy=round(((scores/mean_happy)-1)*100,2))%>%
  mutate(delta_to_sad=round(((scores/mean_sad)-1)*100,2))
print(BW_avg[c(1,4,5)])

```
\
The above chart shows that Finnish people are 1.22% happier than the rest of the "happiest" people and that Afghans are 58.53% less happy than the "happiest" people. We can also see that Finnish people are 136.69% happier than the "least happy" people and Afghans are 3.03% less happy than the "least happy" people. 
\
\
Let's now explore the average happiness ratings across the years by region of the globe instead of by country. 
\
\
```{r echo=F}
##create list of the regions in the data frame and find the average happiness scores for each
regions<-unique(df$Region)
scores2<-c()
for (x in regions) {
  new_df2<-df%>%
    filter(Region==x)
  mean2<-mean(new_df2$Happiness.Score,na.rm=T)
  scores2<-c(scores2,mean2)
}

```
```{r echo=F}
##put into new data frame and rank the regions by the happiness score
df_avg2<-data.frame(regions, scores2)
df_avg2<-df_avg2%>%
  arrange(desc(scores2))

##show top and bottom 4 regions and re-index for logical sense
BW_reg_avg<-rbind(head(df_avg2,4),tail(df_avg2,4))
rownames(BW_reg_avg)<-c(1,2,3,4,-4,-3,-2,-1)
print(BW_reg_avg[1])
```
\
\
Surprisingly, this does not list Western Europe, which is home to the five happiest countries, as number one on our list of "happiest" regions. This implies that the rest of Western Europe is not as "happy" as these countries. Perhaps Western Europe should have been separated out into Northern and Western Europe.
It is not surprising that Sub-Saharan Africa is listed as the "least happy" region, since it is home to four of the "least happy" countries. 
\
\
We can now look at a plot to visualize the difference in scores across these regions. 
\
```{r echo=F, warning=F}
par(mrow=c(1,2))
ggplot(data=head(BW_reg_avg,4),aes(x=regions, y=scores2))+
  geom_point()+
  labs(x='Region', y='Happiness Score', title='Happiest Regions')
ggplot(data=tail(BW_reg_avg,4),aes(x=regions, y=scores2))+
  geom_point()+
  labs(x='Region', y='Happiness Score', title='Least Happy Regions')
par(mrow=c(1,1))
```
\
In the above plots, it shows that Austaliza and New Zealand and North America are significantly "happier" than the other "happiest" regions. Similarly, the Middle East and North Africa seems to be significantly "happier" than the other "least happy" regions.
\
\
Now, let's look at the deltas for the regions, as we did for the individual countries.
\
\
```{r echo=F}
##Figure the percent difference between the happiest/least happy regions and the average happiness of all regions
mean_score2<-mean(df_avg2$scores2)
BW_reg_avg<-BW_reg_avg%>%
  mutate(delta_to_mean2=round(((scores2/mean_score2)-1)*100,2))
print(BW_reg_avg[c(1,3)])
```
\
The above chart shows that Australia and New Zealand are 28.39% happier than other regions of the world and Sub-Saharan Africa is 24.17% less happy than other regions. 
\
\
Let us also compare each region to the average score for the four "happiest" regions and average score for the four "least happy" regions.
\
```{r echo=F}
#Identify mean scores for the happiest regions and mean scores for the least happy regions
mean_happy2<-mean(head(BW_reg_avg,4)$scores2)
mean_sad2<-mean(tail(BW_reg_avg,4)$scores2)

#find the percent differences 
BW_reg_avg<-BW_reg_avg%>%
  mutate(delta_to_happy2=round(((scores2/mean_happy2)-1)*100,2))%>%
  mutate(delta_to_sad2=round(((scores2/mean_sad2)-1)*100,2))
print(BW_reg_avg[c(1,4,5)])

```
\
The above chart shows that Australia and New Zealand are 3.81% happier than the "happiest" regions and Sub-Saharan Africa is 38.69% less happy than the "happiest" regions. We can also see that Australia and New Zealand are 57.29% happier than the "least happy" regions and Sub-Saharan Africa is 7.10% less happy than the "least happy" regions. 
\
\
We can conclude from this analysis that to be in the "happiest" region, it does not necessarily mean one is in the "happiest" nation. The same can be said for the "least happy" region and nation. 